### 2012-06-30 2.tex
Feverish writing on chapters 2-4

\citet{Halle1962} and \citet{Chomsky1965} observe that speakers internalize generalizations about possible and impossible words in their language. \citeauthor{Chomsky1965} illustrate this knowledge with two nonce words \emph{blick} [blɪk] and \emph{bnick} [bnɪk]; whereas \emph{blick} is unattested, but possible word of English, 
speakers immediately recognize \emph{bnick} to be ill-formed, an impossible word of the language.

The classic account of this contrast is as follows. Segments must be assigned to prosodic structures like the syllable, or undergo phonological repair (e.g., \citealt{Hooper1973}:10f., \citealt{Ito1989a}, \citealt{Noske1992}). In English, [bl] is a permissible onset, but unlike some languages (e.g., Moroccan Arabic \emph{bni} `building', \emph{bnat} `daughters', \emph{bniqa} `closet'), [bn] is not. \citet[][19f.]{Wolf2009} suggest, for instance, that an underlying /bnɪk/ would be realized as [nɪk]; it is also possible to imagine resolution by prothesis---[əbnɪk]---or anapytxis---[bənɪk], and indeed \citet{Davidson2006b} finds that English speakers produce these repairs when attempting to mimic foreign pronunciations of obstruent-nasal onsets.

Many linguists view this account as overly simpliest, for, as \citeauthor{Shademan2006} argues, it draws a categorical boundary between the possible and impossible which fails to match the granularity at which speakers may report wordlikeness intuitions.

@@

This chapter focuses on the hypothesis that speakers' knowledge of wordlikeness is gradient. \S\ref{gradhyp} places this debate in a historical perspective, and argues that the naïve but widely accepted form of this hypothesis fails the test of falsifiability. In \S\ref{2evaluation}, a revision of this hypothesis is subjected to an quantitative evaluation and is found wanting.

[Explain that there's not going to be any issue of, e.g., heterovoiced obstruents, that stuff is all in chap. 4]

@@

%In this section, the gradience hypothesis is reviewed from historical and scientific perspectives.

@@

\citet{Chomsky1965} were not the first to consider contrasts between possible and impossible words. Their primary contribution to this question, is, essentially, mentalism: they recognize that naïve speakers effortlessly acquire language-specific wordlikeness generalizations and can report them with ease. But the nature of these contrasts were also considered by structuralists. In an early study, \citet{Fischer-Jorgensen1952} proposes that possible and impossible words represent endpoints on a continuum, and that no non-arbitrary line can be drawn between them. While it would anachronistic to assign this a mentalistic interpretation, this appears to be the earliest proposal that wordlikeness contrasts are in some sense gradient.

@@

Gradience is also discussed in the earliest work in generative syntax. \citet[][132]{LSLT} writes that ``there is little doubt that speakers can fairly consistently order new utterances, never previously heard, with respect to their degree of `belongingness' to the language''. \citet{ASPECTS} proposes to link different degrees of ungrammaticality to different kinds of syntactic violations. Unlike \citeauthor{Fischer-Jorgensen1952}, there is for \citeauthor{LSLT} a hard-and-fast boundary between the grammatical and ungrammatical, only the latter showing any gradience. For \citeauthor{LSLT}, grammatical utterances are all equally grammatical, but ungrammatical utterances may be deviant in their their own ways \citep[][61]{Schutze1996}.

However, most of the early literature on wordlikeness concerns apparently categorical contrasts the possible and impossible. \citet[][31]{Vogt1954} recognizes that the taxonomic phoneme is insufficient to account for many wordlikeness contrasts. Generalizing somewhat from \citeauthor{Vogt1954}'s discussion, allophony may account for the absence of certain phone sequences, but it does not provide a suitable explanation for the absence of initial [bn] in English, nor does it make correct predictions about the surface realization of an underlying initial /bn/. \citeauthor{Vogt1954} concludes that additional grammatical machinery will be needed to account for possible and impossible words. \citet{Halle1962} and \citet{Stanley1967} propose to derive wordlikeness effects from morphophonemic redundancy rules acting on underlying representations. The discussion of \emph{bnick} by \citet[][101]{Chomsky1965} is a good example of this type of analysis. They correctly observe that before a word-initial stop, a English consonant must be liquid. A side effect of a redundancy rule which specifies a consonant in this position as [$+$\textsc{Liquid}] precludes the derivation of \emph{bnick}.\footnote{This is not the only redundancy that might be used to rule out \emph{bnick}; for instance, the only word-initial obstruent that may follow a nasal in English is /s/.}

@@

In \emph{The sound pattern of English} (SPE), \citet{SPE} extend this model of wordlikeness to account for

[Explain the SPE model]

@@

The \emph{SPE} model was not widely adopted, and research on wordlikeness turned to other principles. Syllabification plays no role in \emph{SPE} theory, though the syllable is found in many earlier studies (see \citealp{Goldsmith2011b}). \citet{Hooper1973} and \citet{Kahn1976} argue that wordlikeness generalizations argue for the need for syllabification as part of the phonological computation. This classic syllabification-based model is further enriched by the autosegmental theory of the syllable \citep{McCarthy1979b}, which envisions the syllable as an articulated tree structure \citep[as envisioned by][]{Pike1947a}, and the theory of prosodic licensing \citep{Ito1989a}, in which syllabification triggers phonological repairs.

@@

The observation itself cannot be questioned, but \citeauthor{Hayes2008a} and \citeauthor{Albright2009a} do not explicitly state why this data is relevant to the construction of models of wordlikeness. The following seems the most likely: \citeauthor{Hayes2008a} and \citeauthor{Albright2009a} believe this proves that wordlikeness, as an internal state, is gradient simply because subjects make use of intermediate degrees of wordlikeness in judgement tasks. This proposition, generalized below, is ``naïve'' not because it lacks sophistication, but because it is rooted in a belief in naïve realism, a philosophy which holds that perception provides a relatively direct picture of the nature of the world, and an influential view in the cognitive sciences (see \citealt{Fodor1981a} for a critique).

@@

\begin{quote}
Are there definitional concepts? Of course. For example. consider the concept \emph{odd number}. This seems to have a clear definition, a precise description\ldots{}No integer seems to sit on the fence, undecided as to whether it is quite even, or perhaps a bit odd. No odd number seems odder than any other odd number. \citep[274]{Armstrong1983}
\end{quote}

\citeauthor{Armstrong1983} note that if subjects produce intermediate ratings for these categories, then the naïve proposition is false,
%(i.e., $p \wedge \neg q \vdash \neg (p rightarrow q)$
and this is what \citeauthor{Armstrong1983} find in their experiments. They ask subjects to rate, on a seven-point scale, the extent to which, e.g., certain odd counting numbers represent the concept ``odd number'', and find that speakers do make use of intermediate values. For instance, \emph{7} is rated 1.4 out of 7 on average, whereas the average ratings of \emph{447} and \emph{501} are 3.5 and 3.7, respectively.

In summary, the proposition that intermediate ratings indicate mental gradience is either untestable, or it is false, and is thus rejected.

@@

\citeauthor{Armstrong1983} do not view their results as evidence that subjects have an internal model of odd numbers which is at odds with the formal, all-or-nothing definition, and in fact, several additional experiments show that subjects deploy the extension of the formal definition in other tasks. Reviewing this study, \citet[215]{Schutze2011} writes that the experiments show that judgements are ``sensitive to factors other than our underlying competence''. In the case of wordlikeness, a case can be made that one such factor is similarity to existing words (e.g., \citealt[][151, fn. 27]{LSLT}, \citealt{Greenberg1964}, \citealt{Ohala1986b}, \citealt{Bailey2001}). However, the mere fact that subjects use intermediate ratings does not show that they do so with any systematicity, or that, e.g., the contrast between \emph{447} and \emph{7} with respect could be given any satisfying explanation. \citeauthor{Schutze2011} further suggests that gradience responses is a property of tasks, not concepts.

\begin{quote}
\ldots{}when asked for gradient responses, participants will find some way to oblige the experimenter; if doing so is incompatible with the experimenter's actual question, they apparently infer that she must have really intended to ask something slightly different. \citep[][215]{Schutze2011}
\end{quote}

@@

The remainder of this chapter is devoted to evaluting the gradience hypothesis, using state-of-the-art computational wordlikeness models and data from English.

@@

The evaluation makes exclusive use of previously published English wordlikeness data. For inclusion in the evaluation, a study must conform to all three of the following conditions. First, the stimuli must consist only of monosyllabic words presented auditorily. Secondly, a significant portion of the stimuli must contain gross phonotactic violations (e.g., \emph{bnick}). Finally, the ratings, averaged across subjects, must be publicly available. The large-scale studies by \citet{Bailey2001} and \citet{Shademan2006,Shademan2007} are ineligible because the former lack stimuli with gross phonotactic violations and the latter data are not available to the public in any form.

@@

\citet{Greenberg1964} investigated wordlikeness using the technique of free magnitude estimation, a mechanism which has become increasingly popular among syntactians \citep[e.g.,][]{Bard1996}. At the beginning of the experiment, the subject heard a recording of the word \emph{stick}. In subsequent trials, the subjects heard a nonce word and were asked to report ``how far would you say that is from English?'', with \emph{stick} at ``1''; subjects are told that a word that is ``twice as far from English'' as \emph{stick} should be scored ``2''. The data used here are from \citeauthor{Greenberg1964}'s Experiment B, in which 17 undergraduates were presented 17 stimuli in all. In addition to \emph{stick}, the stimuli include three other English words; these four items were excluded from further analyses, leaving 13 stimuli. As is standard practice in psychophysics \citep[e.g.,][]{Butler1987}, these magnitude ratings were log-transformed before analysis.

@@

\citet{Scholes1966} conducted a number of English wordlikeness judgement with middle school children. The data used here come from his ``experiment 5'', in which 63 monosyllabic items were presented to 33 seventh-grade students. For each stimulus, the subjects produced forced choice ``yes''/``no'' answers to the question of whether the item ``is likely to be usable as a word of English''. \citet{Hayes2008a} and \citet{Albright2009a} analyse this data as gradient by performing an item averaging using the fraction of ``yes'' answers for each stimulus (see also \citealt{Pierrehumbert1994}, \citealt{Coleman1997}, \citealt{Frisch2000}). For instance, 22 of the 33 students answered ``yes'' for \emph{shlerk} [ʃlɚk], so it is assigned a score of $0.666$.\footnote{This procedure conflates intraspeaker variation (which may be considerable; see \citealt{Shademan2007}) and speaker-internal gradience, and its adoption here should not be construed as an endorsement. It is provided solely for comparison with prior studies using the \citeauthor{Scholes1966} data.} These stimuli are all ostensibly non-words, but include \emph{clung} [klʌŋ], the preterite and past participle of the verb \emph{cling}, and \emph{brung} [brʌŋ], a dialectical past participle for \emph{bring}. These two words were excluded and the remaining 61 stimuli were submitted to analysis.

@@

The four models used here consist of a binary baseline and three computationally implemented gradient models. These three models are chosen because prior studies have shown they are correlated with wordlikeness judgements. Non-parametric rank correlation statistics are used evaluate the correlation between model predictions and wordlikeness ratings. Rank correlation are used rather than the parametric Pearson correlation statistic that is used by \citet{Hayes2008a}, for instance, because they make none of the potentially troublesome assumptions of the latter method (see \citealt{Albright2009a}:23, fn. 12). The Spearman $\rho$ is most widely known rank correlation statistic, but it is difficult to give a natural interpretation to this quantity. On the other hand, the Kendall $\tau_b$ and Goodman-Kruskal $\gamma$ can be interpreted as fractions of the number of \emph{concordant} and \emph{discordant} pairs \citep{Noether1981}. Consider the case here, in which model scores are compared with wordlikeness ratings. If a model rates \emph{dresp} [dɹɛsp] more wordlike than *\emph{srest} [sɹɛst], this pair is concordant if speakers also rate \emph{dresp} more English-like than \emph{srest}, and it is discordant if *\emph{srest} is rated more English-like. The $\tau_b$ and $\gamma$ differ only in their treatment of ``ties'' (i.e., if e.g., *\emph{dresp} and *\emph{srest} are scored the same, or rated the same); $\tau_b$ includes a correction for ties, whereas $\gamma$ ignores tied pairs. Much like the familiar Pearson correlation, $\rho$, $\tau_b$ and $\gamma$ are all in the range [-1, 1]. Correlations of the four models are given in Table \ref{cor}.

\begin{table} \label{cor}
\centering
\begin{tabular}{l l l l l}
\toprule
Spearman $\rho$          & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {0.845}          & {0.765} & {\textbf{0.863}} & {0.648}          \\
\citealt{Scholes1966}    & {0.791}          & {0.762} & {\textbf{0.827}} & {0.827}          \\
\citealt{Albright2003b}  & {0.725}          & {0.429} & {0.708}          & {\textbf{0.742}} \\
\midrule
Kendall $\tau_b$         & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {\textbf{0.716}} & {0.585} & {0.692}          & {0.462}          \\
\citealt{Scholes1966}    & {\textbf{0.664}} & {0.597} & {0.652}          & {0.565}          \\
\citealt{Albright2003b}  & {\textbf{0.599}} & {0.343} & {0.506}          & {0.556}          \\
\midrule
Goodman-Kruskal $\gamma$ & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {\textbf{1.000}} & {0.684} & {0.692}          & {0.462}          \\
\citealt{Scholes1966}    & {\textbf{0.995}} & {0.634} & {0.667}          & {0.614}          \\
\citealt{Albright2003b}  & {\textbf{0.953}} & {0.656} & {0.509}          & {0.575}          \\
\bottomrule
\end{tabular}
\caption{Rank correlations between wordlikeness ratings and phonotactic models surprisingly reveal that the binary baseline meeds or exceeds the coverage of three state-of-the-art phonotactic models. All correlations are significant at $p = 0.05$.}
\end{table}

@@

%As \citet{Halle1962} argues, this eliminates the need for additional grammatical machinery.
%\begin{quote}

>>>>

\citet{Halle1962} and \citet{Chomsky1965} observe that speakers internalize generalizations about possible and impossible words in their language. \citeauthor{Chomsky1965} illustrate this knowledge with two nonce words \emph{blick} [blɪk] and \emph{bnick} [bnɪk]. Whereas \emph{blick} is unattested, but in some sense a ``possible'' word of English, speakers immediately recognize \emph{bnick} to be ill-formed, an impossible word of the language.

A naïve account of this contrast is based on the assumption that segments or timing units must either be parsed into prosodic structures like the syllable, or be subject to phonological repair. Early precedents for this can be found in dissertations by \citet[10f.]{Hooper1973} and \citet[57f.]{Kahn1976}, and it is a fundamental component of many later frameworks (e.g., \citealt{Ito1989a}, \citealt{Noske1992}, \citealt{OT}). In English, [bl] is a permissible onset, but unlike some languages (e.g., Moroccan Arabic \emph{bni} `building', \emph{bnat} `daughters', \emph{bniqa} `closet'), [bn] is not. \citet[][19f.]{Wolf2009} suggest, for instance, that an underlying /bnɪk/ would be realized as [nɪk]; it is also possible to imagine resolution by prothesis---[əbnɪk]---or anapytxis---[bənɪk], and indeed \citet{Davidson2006b} finds that English speakers use all three of these repairs when asked to mimic foreign pronunciations of obstruent-nasal onsets like /bn/.

Many linguists who have concerned themselves with ``wordlikeness'' now regard the above as overly simplistic, for, as \citeauthor{Shademan2006} argues, such accounts draw a categorical boundary between the possible and impossible which fails to match the granularity at which speakers may report ``wordlikeness'' intuitions.

@@

This chapter focuses on the hypothesis, implicit in \citeauthor{Shademan2006}'s statement, that there is an inherent gradient or continuum between possible and impossible words, against an null hypothesis that the observed patterns of judgements are little more than an artifact of the wordlikeness task itself. \S\ref{gradhyp} places this debate in a historical perspective, and argues that the naïve but widely accepted form of this hypothesis fails the test of falsifiability. In \S\ref{2evaluation}, a quantitative evaluation of a revised version of this hypothesis reveals a poor match between the gradience in speakers' judgements and contemporary computational models of wordlikeness. Taken together, the findings argue against the gradience hypothesis.

% What \citet[32]{Cohn2006} calls ``gradient well-formedness''

@@

@@

\citeauthor{Chomsky1965} were not the first to consider the notion of possible and impossible words. Their primary contribution is that their mentalist perspective: they recognize that naïve speakers effortlessly acquire language-specific generalizations about possible and impossible words and can report them without any explicit training.

\subsubsection{The Copenhagen school}

Wordlikeness received some attention by structuralists of the Copenhagen school. In an early study, \citet{Fischer-Jorgensen1952} proposes that possible and impossible words represent endpoints on a continuum, and that no non-arbitrary line can be drawn between them. While it would anachronistic to assign this a mentalistic interpretation, this appears to be the earliest proposal that wordlikeness contrasts are in some sense gradient.

However, not all early literature is concerned with gradience. \citet[][31]{Vogt1954}, for instance, recognizes that the taxonomic phoneme is insufficient to account for many wordlikeness contrasts. \citeauthor{Vogt1954} observes that allophony may account for the absence of certain phone sequences, but it does not provide a suitable explanation for the absence of initial [bn] in English, nor does it make correct predictions about the surface realization of an underlying initial /bn/. \citeauthor{Vogt1954} concludes that additional grammatical machinery will be needed to account for possible and impossible words.

@@

Gradient grammaticality is assumed in the earliest work on generative syntax, \emph{The logical structure of linguistic theory}, in which \citet[][132]{LSLT} writes that ``there is little doubt that speakers can fairly consistently order new utterances, never previously heard, with respect to their degree of `belongingness' to the language''. Later, \citet{ASPECTS} proposes that different syntactic violations give rise to different degrees of ungrammaticality. Unlike \citeauthor{Fischer-Jorgensen1952}, however, for \citeauthor{LSLT} this continuum does incorporate a hard-and-fast boundary between the grammatical and ungrammatical: there are no constrasts between grammatical utterances, merely between different degrees of ungrammaticality. For \citeauthor{LSLT}, all grammatical utterances are alike with respect to their grammaticality, but there are many degrees of ungrammaticality \citep[][61]{Schutze1996}.

\citet{Halle1962} and \citet{Stanley1967} concern themselves only with categorical wordlikeness contrasts, however. Both authors derive wordlikeness effects from morphophonemic redundancy rules acting on underlying representations. The discussion of \emph{bnick} by \citet[][101]{Chomsky1965} is a good example of this type of analysis. \citeauthor{Chomsky1965} observe that a consonant before a word-initial stop can only be a liquid in English. A corresponding redundancy rule which specifies a consonant in this position as [$+$\textsc{Liquid}] precludes the derivation of \emph{bnick}, since there is no way to realize the second consonant as a nasal.\footnote{This is not the only redundancy that could rule out \emph{bnick}. For instance, /s/ is the only word-initial obstruent that may precede a nasal consonant in English.}

@@

In \emph{The sound pattern of English} (SPE), \citet{SPE} argue that a complete solution to wordlikeness must countenance gradience. In addition to the contrast of \emph{blick} and \emph{bnick}, they introduce a third nonce word \emph{bnzk}, which they take to be even less well-formed than ``impossible'' \emph{bnick}.

\begin{quote}
Hence, a real solution to the problem of ``admissibility'' will not simply define a tripartite categorization of occurring, accidental gap, and inadmissible, but will define the `degree of admissibility' of each potential lexical matrix in such a way as to\ldots{}make numerous other distinctions of this sort (\emph{SPE}:416--417)
\end{quote}

In some languages (e.g., Imdlawn Tashlhiyt Berber [tm.zħ] `she jested', [tzmt] `it (f.) is stifling'; \citealt{Dell1985}), \emph{bnzk} is syllabifiable, so its ill-formedness is once again a property of English, not a universal.

\citeauthor{SPE} propose to derive gradience from the complexity (using a simple feature-counting metric) of the redundancy rules needed to map a nonce word to an attested word, and they show that this derives the increasing cline of wordlikeness running from ``consonant soup'' \emph{bnzk} to ill-formed \emph{bnick}, possible \emph{blick}, and finally lexical \emph{brick}.

Despite the considerable attention given to the proposals of \emph{SPE} in the wake of that book's publication in 1968, the \emph{SPE} wordlikeness model has received almost no attention in the literature. At the risk of explaining what might be no more than an an accidental gap in the historical record, the novel aspects of the \emph{SPE} model---gradience derived from similarity to existing lexical entries---seem to have been overshadowed by compelling arguments against \citeauthor{SPE}'s assumption (see also \citealt{Halle1962}, \citealt{Stanley1967}) that wordlikeness contrasts derive from properties of underlying forms. \citet{Shibatani1973} observes that there are some generalizations about surface forms which give rise to wordlikeness contrasts, but cannot be stated as constraints on underlying forms. An example from German is shown in (\ref{fd}) below.

\begin{example}[German final devoicing] \label{fd}
\begin{tabular}{l l l l}
   & nom.sg. & nom.pl.    \\
a. & [piːp]    & [piːpə]  & `cheep(s)'      \\
   & [diːp]    & [diːbə]  & `thief/thieves' \\
b. & [ɡʀaːt]   & [ɡʀaːtə] & `ridge(s)'      \\
   & [ɡʀaːt]   & [ɡʀaːdə] & `degree(s)'     \\
\end{tabular}
\end{example}

\noindent The plural appears to preserve a contrast in final obstruent voicing which is absent in the singular:\footnote{While there is a contentious debate as to whether devoicing in German is completely neutralizing (e.g., \citealt{Fourakis1984}) or not \citep[e.g.,][]{Port1985}, it is irrelevant to the discussion at hand.} word-final voiced stops are never found in German, and \citet[95]{Shibatani1973} reports that ``it is easy to show that a native speaker of German rejects those forms \emph{on the ground that they end in voiced obstruents}'' (emphasis in original). However, given that root-final voicing is not predictable (e.g., /ɡʀaːt \alt{} ɡʀaːd/), the process of \textsc{Final Devoicing} cannot be any kind of lexical redundancy and is mysterious under the \emph{SPE} account. The alternative proposed by \citeauthor{Shibatani1973} and by \citet{Clayton1976}, is that to say that nonce words like [ɡʀaːd] are ill-formed not beacuse of any underlying property, but because they fail to undergo an otherwise-exceptionless phonological process, \textsc{Final Devoicing}, or equivalently, the surface-true generalization it derives.

\citet{Sommerstein1974} further notes that the computation of hypothetical underlying forms from nonce surface forms which is implied by the \emph{SPE} theory is non-trivial in the presence of phonological opacity \citep[see][528f.]{Anderson1988a}, and the rejection of the biuniqueness principle means that there is not always a unique solution, as is illustrated by the two underlying forms corresponding to [ɡʀaːt] in (\ref{fd}) above.

@@

The arguments of \citeauthor{Shibatani1973} and others led theorists to focus their attention on properties of surface representations as determinants of wordlikeness. Though syllabification plays no role in \emph{SPE}, the syllable is discussed in many earlier studies, and it received particular attention in the 1970s. \citet{Hooper1973} and \citet{Kahn1976} argue that the syllable is useful for defining wordlikeness generalizations.\footnote{\citet{Steriade1999} and \citet{Blevins2003}, however, argue that a number of phonotactic generalizations previously stated in syllabic terms can be reanalyzed without making reference to syllables.} \citeauthor{Hooper1973} argues, for instance, that [bn], impossible as an English onset, is unobjectionable as a syllable contact cluster in nonce words like \emph{stabnik} (or in names like \emph{Abner}), and that this demonstrates the superiority of syllable-based wordlikeness generalizations. This already signals further trouble for alternative accounts which focus on underlying forms. Syllabification may span morphs, is generally predictable, and is universally non-contrastive, and as a consequence, few posit in to be present in underlying representations \citep[though see, e.g.,][]{Vaux2003}. \citeauthor{Hooper1973} also points to loanword adaptations which produce native syllable structure \citep[e.g.,][]{Carlisle1991} as evidence that syllabification is part of the phonological computation.

Further enrichments to the theory are provided by the autosegmental theory of the syllable \citep{McCarthy1979b}, which envisions the syllable as an articulated tree structure \citep[as envisioned by][]{Pike1947a}, and theories like prosodic licensing \citep{Ito1989a}, in which syllabification triggers phonological repairs.

While none of these authors discuss gradience, the syllable plays a role in the definition of a gradient measure, ``positional probability'', though to correlate closely with human judgements of wordlikeness. The positional probability of a nonce monosyllabic word is derived from the combined probabilities at which segments occur in onset, nucleus, and coda positions in the lexicon of a given language.

Using the head-term preference paradigm, \citet{Jusczyk1993b} and \citet{Friederici1993} find that typically-developing children as young as 9 months of age distinguish between nonce words which are and are not phonotactically valid in their target language. \citet{Jusczyk1994} report that 9-month-old children acquiring English also show preferences for nonce words with high positional probability over those with low positional probability. Faciliatory effects of positional probability (i.e., shorter latencies) are reported for other nonce word tasks conducted with adults, including single-word shadowing \citep{Vitevitch1997,Vitevitch1998}, same/different judgements (\citealt{Vitevitch1999a}, \citealt{Luce2001}, \citealt{Lipinski2005}, \citealt{Vitevitch2005}), and lexical decision \citep{Pylkkanen2002a}.

%\citet{Mattys1999,Mattys2001}
%\citet{McQueen1998}

The most interesting result of these studies comes from the first experiment of \citealt{Vitevitch1997}, which finds that speakers' wordlikeness ratings of bisyllabic words are correlated with the positional probabilities of the constituent syllables. Unfortunately, neither \citeauthor{Vitevitch1997} nor any of the other aforementioned psycholinguists make any effort to eliminate the possibility that the low positional probability stimuli are ``impossible'' words of English. In fact, in Chapter \ref{clusters}, it is argued that many of the stimuli used by \citet{Vitevitch1997}, including even the high probability stimuli, contain illicit word-medial consonant clusters. While \citeauthor{Vitevitch1997} neither control nor manipulate the well-formedness of medial clusters, in a post-hoc test they consider a probabilistic measure of cluster well-formedness to a post-hoc test, which reveals that cluster well-formedness is correlated with syllable-internal positional probabilities and the wordlikeness judgements. \citeauthor{Vitevitch1997} ultimately conclude that it cannot explain all the variation in wordlikeness.

The aforementioned studies all conclude that the gradient measure of positional probability correlates with behavioral results. As the flaws of the \citet{Vitevitch1997} study demonstrate, the aforementioned studies do little to tease apart the gradient and categorical aspects of phonotactics. More generally, they do little to distinguish between positional probability and closely correlated measures like bigram probability (see \S\ref{bigram} below) or neighborhood density, since these studies carefully select stimuli which either have high or low values for all of positional probability, bigram probablility, and neighborhood density. This is particularly troublesome given that positional probability was created \emph{ex nihilo}; in contrast, the effects of neighborhood density in various psycholinguistic tasks are emergent properties of many models of speech production (e.g., \citealt{Luce1998}, \citealt{Luce2000}) and perception (e.g., \citealt{Marslen-Wilson1984,Marslen-Wilson1987}, \citealt{McClelland1986}, \citealt{Norris1994}, \citealt{Norris2000}).

@@

The observation itself cannot be questioned, but \citeauthor{Hayes2008a} and \citeauthor{Albright2009a} do not explicitly state why this data is relevant to the construction of models of wordlikeness. Presumably, these authors believe that these patterns of judgements demonstrate that wordlikeness, as an internal state, is gradient simply because subjects make use of intermediate degrees of wordlikeness in judgement tasks. This proposition, generalized below, is ``naïve'' not because it lacks sophistication, but because it is rooted in a belief in naïve realism, a philosophy which holds that perception provides a relatively direct picture of the nature of the world, and in general, an influential view in the cognitive sciences (see \citealt{Fodor1981a} for a critique).

@@

\begin{quote}
Are there definitional concepts? Of course. For example, consider the concept \emph{odd number}. This seems to have a clear definition, a precise description\ldots{}No integer seems to sit on the fence, undecided as to whether it is quite even, or perhaps a bit odd. No odd number seems odder than any other odd number. \citep[274]{Armstrong1983}
\end{quote}

\citeauthor{Armstrong1983} note that if subjects produce intermediate ratings for these categories, then the naïve proposition is false, and this is what \citeauthor{Armstrong1983} find in their experiments. They ask subjects to rate, on a seven-point scale, the extent to which, e.g., certain odd counting numbers represent the concept ``odd number'', and find that speakers do make use of intermediate values, as can be seen in Figure \ref{agg}. In summary, the proposition that intermediate ratings indicate mental gradience is either untestable, or it is false, and is thus rejected.

\begin{figure} \centering
\includegraphics{agg.pdf}
\caption{Subjects asked to rate the degree to which certain even and odd numbers represent the categories ``even'' and ``odd'', respectively, freely use intermediate ratings.}
\label{agg} 
\end{figure}

@@

\citeauthor{Armstrong1983} do not view their results as evidence that subjects have an internal model of odd numbers which is at odds with the formal, all-or-nothing definition, and in fact, several additional experiments show that subjects deploy the extension of the formal definition in other tasks. Reviewing this study, \citet[215]{Schutze2011} writes that the experiments show that judgements are ``sensitive to factors other than our underlying competence''. In the case of wordlikeness, a case can be made that one such factor is similarity to existing words (e.g., \citealt[][151, fn. 27]{LSLT}, \citealt{Greenberg1964}, \citealt{Ohala1986b}, \citealt{Bailey2001}, \citealt{Schutze2005}). However, the mere fact that subjects use intermediate ratings does not show that they do so with any systematicity, or that, e.g., the contrast between \emph{447} and \emph{7} with respect could be given any satisfying explanation. \citeauthor{Schutze2011} further suggests that gradience responses is a property of tasks, not concepts.

\begin{quote}
Putting it another way, when asked for gradient responses, participants will find some way to oblige the experimenter; if doing so is incompatible with the experimenter's actual question, they apparently infer that she must have really intended to ask something slightly different. \citep[][215]{Schutze2011}
\end{quote}

@@

The remainder of this chapter is devoted to evaluating the gradience hypothesis, using state-of-the-art computational wordlikeness models and data from English.

@@

The evaluation makes exclusive use of previously published English wordlikeness data. For inclusion in the evaluation, a study must conform to all three of the following conditions. First, the stimuli must consist only of monosyllabic words presented auditorily. Secondly, a significant portion of the stimuli must contain gross phonotactic violations (e.g., \emph{bnick}). Finally, the ratings, averaged across subjects, must be publicly available. The large-scale studies by \citet{Bailey2001} and \citet{Shademan2006,Shademan2007} are ineligible because the former lack stimuli with gross phonotactic violations and the latter data are not available to the public in any form. All data are reproduced in Appendix \ref{appendixA}

@@

\citet{Greenberg1964} investigated wordlikeness using the technique of free magnitude estimation, a mechanism which has become increasingly popular among syntacticians \citep[e.g.,][]{Bard1996}. At the beginning of the experiment, the subject heard a recording of the word \emph{stick}. In subsequent trials, the subjects heard a nonce word and were asked to report ``how far would you say that is from English?'', with \emph{stick} at ``1''; subjects are told that a word that is ``twice as far from English'' as \emph{stick} should be scored ``2''. The data used here are from \citeauthor{Greenberg1964}'s Experiment B, in which 17 undergraduates were presented 17 stimuli in all. In addition to \emph{stick}, the stimuli include three other English words; these four items were excluded from further analyses, leaving 13 stimuli. As is standard practice in psychophysics \citep[e.g.,][]{Butler1987}, these magnitude ratings were log-transformed before analysis.

@@

\citet{Scholes1966} conducted a number of English wordlikeness judgement with middle school children. The data used here come from his ``experiment 5'', in which 63 monosyllabic items were presented to 33 seventh-grade students. For each stimulus, the subjects produced forced choice ``yes''/``no'' answers to the question of whether the item ``is likely to be usable as a word of English''. \citet{Hayes2008a} and \citet{Albright2009a} analyze this data as gradient by performing an item averaging using the fraction of ``yes'' answers for each stimulus (see also \citealt{Pierrehumbert1994}, \citealt{Coleman1997}, \citealt{Frisch2000}). For instance, 22 of the 33 students answered ``yes'' for \emph{shlerk} [ʃlɚk], so it is assigned a score of $0.666$.\footnote{This procedure conflates intraspeaker variation, which in some cases is considerable \citep{Shademan2007} with speaker-internal gradience, and its adoption here should not be construed as an endorsement. Rather, it is provided solely for comparison with prior studies using the \citeauthor{Scholes1966} data.} These stimuli are all ostensibly non-words, but include \emph{clung} [klʌŋ], the preterite and past participle of the verb \emph{cling}, and \emph{brung} [brʌŋ], a dialectical past participle for \emph{bring}. These two words were excluded and the remaining 61 stimuli were submitted to analysis.

@@

The four models used here consist of a binary baseline and three computationally implemented gradient models. These three models are chosen because prior studies have shown they are correlated with wordlikeness judgements. Non-parametric rank correlation statistics are used evaluate the correlation between model predictions and wordlikeness ratings. Rank correlation are used rather than the parametric Pearson correlation statistic that is used by \citet{Hayes2008a}, for instance, because they make none of the potentially troublesome assumptions of the latter method (see \citealt{Albright2009a}:23, fn. 12). The Spearman $\rho$ is most widely known rank correlation statistic, but it is difficult to give a natural interpretation to this quantity. On the other hand, the Kendall $\tau_b$ and Goodman-Kruskal $\gamma$ can be interpreted as fractions of the number of \emph{concordant} and \emph{discordant} pairs \citep{Noether1981}. Consider the case here, in which model scores are compared with wordlikeness ratings. If a model rates \emph{dresp} [dɹɛsp] more wordlike than *\emph{srest} [sɹɛst] and if speakers further rate \emph{dresp} more English-like than \emph{srest}, than \emph{dresp}/\emph{srest} are a concordant pair. If, however, a model and speakers disagree on the relative ranking of \emph{dresp} and \emph{srest}, the pair is discordant. The $\tau_b$ and $\gamma$ differ only in their treatment of ``ties'' (i.e., if e.g., *\emph{dresp} and *\emph{srest} are scored the same, or rated the same); $\tau_b$ includes a correction for ties, whereas $\gamma$ ignores tied pairs. Much like the familiar Pearson correlation, $\rho$, $\tau_b$ and $\gamma$ are all in the range [$-1$, $1$]. Correlations of the four models are given in Table \ref{cor}.

\begin{table} \centering
\begin{tabular}{l l l l l}
\toprule
Spearman $\rho$          & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {0.845}          & {0.765} & {\textbf{0.863}} & {0.648}          \\
\citealt{Scholes1966}    & {0.791}          & {0.762} & {\textbf{0.827}} & {0.827}          \\
\citealt{Albright2003b}  & {0.725}          & {0.429} & {0.708}          & {\textbf{0.742}} \\
\midrule
Kendall $\tau_b$         & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {\textbf{0.716}} & {0.585} & {0.692}          & {0.462}          \\
\citealt{Scholes1966}    & {\textbf{0.664}} & {0.597} & {0.652}          & {0.565}          \\
\citealt{Albright2003b}  & {\textbf{0.599}} & {0.343} & {0.506}          & {0.556}          \\
\midrule
Goodman-Kruskal $\gamma$ & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {\textbf{1.000}} & {0.684} & {0.692}          & {0.462}          \\
\citealt{Scholes1966}    & {\textbf{0.995}} & {0.634} & {0.667}          & {0.614}          \\
\citealt{Albright2003b}  & {\textbf{0.953}} & {0.656} & {0.509}          & {0.575}          \\
\bottomrule
\end{tabular}
\caption{Rank correlations between wordlikeness ratings and phonotactic models surprisingly reveal that the binary baseline meeds or exceeds the coverage of three state-of-the-art phonotactic models. All correlations are significant at $p = 0.05$.}
\label{cor}
\end{table}

@@
